---
title: "Class 09"
author: "Shuyi Wang"
date: "10/30/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Unsupervised Learning Analysis of Cancer Cells

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass. For example radius (i.e. mean of distances from center to points on the perimeter), texture (i.e. standard deviation of gray-scale values), and  smoothness (local variation in radius lengths). Summary information is also provided for each group of cells including diagnosis (i.e. benign (not cancerous) and and malignant (cancerous)).

Import the data.
.
```{r}
url <- "https://bioboot.github.io/bimm143_W18/class-material/WisconsinCancer.csv"
wisc_df <- read.csv(url)
head(wisc_df)
```

So the diagnosis col is the lables
```{r}
table(wisc_df$diagnosis)
```


```{r}
# Convert the features of the data: wisc.data
wisc.data <- as.matrix(wisc_df[3:32], ncol = 30)
wisc.data
```

```{r}
# Set the row names of wisc.data
row.names(wisc.data) <- wisc_df$id
head(wisc.data)
```

setup a separate new vector called diagnosis to be 1 if a diagnosis is malignant ("M") and 0 otherwise. Note that R coerces TRUE to 1 and FALSE to 0.
```{r}
# Create diagnosis vector by completing the missing code
diagnosis <- as.numeric(wisc_df$diagnosis == "M")
sum(diagnosis)
```

# Performing PCA

Q1. How many observations are in this dataset?
```{r}
dim(wisc.data)
length(wisc.data)
```

Q2. How many variables/features in the data are suffixed with _mean?
```{r}
grep("_mean", colnames(wisc.data))
x <- length(grep("_mean", colnames(wisc.data)))
```
There are `r x` mean measurments in this dataset.

```{r}
sum(x, na.rm = FALSE)
```


Q3. How many of the observations have a malignant diagnosis?


```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale = TRUE)
```

```{r}
# Look at summary of results
summary(wisc.pr)
```



Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?


# Interpreting PCA

```{r}
biplot(wisc.pr)
```
Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

Let make a plot of PC1 vs PC2
```{r}
attributes(wisc.pr)
```

```{r}
dim(wisc.pr$x)
```


```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2])
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = diagnosis+1)
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col = diagnosis+1)
```


# Variance explained

```{r}
# Calculate the variance of each principal component by squaring the sdev component 
pr.var <- wisc.pr$sdev^2
# Each principal component by dividing by the total variance explained of all principal components
pve <- pr.var/sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```


```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
# Plot cumulative proportion of variance explained
plot(cumsum(pve), xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```
Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature  concave.points_mean?




# Section 3

```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)
```

We need a distance matrix for hierachical clustering input...
```{r}
data.dist <- dist(data.scaled)
```

Create a hierarchical clustering model 
```{r}
wisc.hclust <- hclust(data.dist, "complete")
plot(wisc.hclust)
```

# Selecting numbers of clusters

```{r}
#cut the tree so that it has 4 clusters
wisc.hclust.clusters <- cutree(wisc.hclust,k=4)

table(wisc.hclust.clusters)
```

compare the cluster membership to the actual diagnoses
```{r}
table(wisc.hclust.clusters, diagnosis)
```

Q12. Can you find a better cluster vs diagnoses match with by cutting into a different number of clusters between 2 and 10?



# Section 4 K-means clustering




# Section 5 Clustering PCA results

Let’s see if PCA improves or degrades the performance of hierarchical clustering.

Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with complete linkage.
```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
d.pr <- dist(wisc.pr$x[, 1:7])
wisc.pr.hclust <- hclust(d.pr, method="complete")
plot(wisc.pr.hclust)
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=4)
table(wisc.pr.hclust.clusters)
```

```{r}
# Compare to actual diagnoses
table(wisc.pr.hclust.clusters, diagnosis)
```
How well do the k-means and hierarchical clustering models you created in previous sections
```{r}
table(wisc.hclust.clusters, diagnosis)
```

# Section 6




# Bonus section: predicting without PCA model

Take new patient data and to out PCA model from
```{r}
## Predicting Malignancy Of New samples
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)

plot(wisc.pr$x[,1:2], col=diagnosis+1)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
```































