---
title: "Class 08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clusterig

Our first example with **kmeans()** function.

```{r}
# Example plot to see how Rmarkdown works
plot(1:10, typ = "l")
```

Back to kmeans

```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```



```{r}
k <- kmeans(x, centers = 2, nstart = 20 )
k
```


Use the kmeans() function setting k to 2 and nstart=20

Inspect/print the results
Q. How many points are in each cluster?
```{r}
k$size
```

Q. What ‘component’ of your result object details
 - cluster size?
   'k$size'
 - cluster assignment/membership?
 - cluster center?
```{r}
k$cluster
```
```{r}
table(k$cluster)
```
```{r}
k$centers
```
 
 
 Plot x colored by the kmeans cluster assignment and
 add cluster centers as blue points
```{r}
plot(x, col = k$cluster)
palette(c("blue", "red"))
points(k$centers, col = "green", pch = 20, cex = 3)
```
 
Q. Repeat for k=3, which has the lower tot.withinss? 

```{r}
k2 <- kmeans(x, centers = 2, nstart = 20)
k3 <- kmeans(x, centers = 3, nstart = 20)
k4 <- kmeans(x, centers = 4, nstart = 20)
k5 <- kmeans(x, centers = 5, nstart = 20)

k2$tot.withinss
k3$tot.withinss
k4$tot.withinss
k5$tot.withinss

plot(c(k2$tot.withinss,k3$tot.withinss, k4$tot.withinss, k5$tot.withinss), typ = "b")
```




## Hierarchical clustering in R 

Let's try out the **hclust()** function for Hierarchical clustering in R. This function need a distance matrix as input!

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```


```{r}
plot(hc)
# Draw a line on the dendrogram
abline(h=8, col="red")

# Cut the tree to yield cuter membership vector
cutree(hc, h=8)
cutree(hc, k=2)
```



### A more "real"example of data clusting

```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")

# Step 2. Plot the data without clustering
plot(x)

# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )

grp.names <- rep(c("red", "blue", "black"), each=50)

palette(c("red", "blue", "black"))
plot(x, col=grp.names, pch=20)

```


Q. Use the dist(), hclust(), plot() and cutree()
        functions to return 2 and 3 clusters
```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)

grp.2 <- cutree(hc, k=2)
grp.3 <-cutree(hc, k=3)

plot(x, col=grp.3, pch=20)
```



Q. How does this compare to your known 'col' groups?

```{r}
# Use the table() function!!
table(grp.3, col)
```



#### Principle component analysis (PCA)

Let's make up some data with quite a few dimensions

```{r}

## Initialize a blank 100 row by 10 column matrix
mydata <- matrix(nrow=100, ncol=10)

## Lets label the rows gene1, gene2 etc. to gene100
rownames(mydata) <- paste("gene", 1:100, sep="")

## Lets label the first 5 columns wt1, wt2, wt3, wt4 and wt5
## and the last 5 ko1, ko2 etc. to ko5 (for "knock-out")
colnames(mydata) <- c( paste("wt", 1:5, sep=""),
 paste("ko", 1:5, sep="") )

## Fill in some fake read counts
for(i in 1:nrow(mydata)) {
 wt.values <- rpois(5, lambda=sample(x=10:1000, size=1))
 ko.values <- rpois(5, lambda=sample(x=10:1000, size=1))

 mydata[i,] <- c(wt.values, ko.values)
}

head(mydata)
```

# Note the prcomp() 
```{r}
head(t(mydata))
```


Now lets try to find "structure"" in this data with **prcomp)**


```{r}
## lets do PCA
pca <- prcomp(t(mydata), scale=TRUE) 
pca
```


Make a PC plot of PC1 vs PC2. For this I use the $x commponent of our output (i.e. pca$x).

```{r}
## A basic PC1 vs PC2 2-D plot
plot(pca$x[,1], pca$x[,2])
 
```

Lets see now ewll our PCs are doing (i.e. now much variance from)
```{r}
## Precent variance is often more informative to look at
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
```

Make a scree plot 
```{r}
barplot(pca.var.per, main="Scree Plot",
 xlab="Principal Component", ylab="Percent Variation")
```


Lets make our PC plot a bit more useful...
```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)")) 
```








